---
title: "Full_Cognition_Script"
author: "Austin Russell"
date: "2025-04-23"
output: html_document
---

```{r}
library(DECIPHER)
library(phyloseq)
library("phangorn")
library(msa)
library(dada2)
library(ape)

#make a tree with MAFFT and FastTree
tree <- read.tree("/Users/russ/Documents/aim_2_paper_cognition/MAFFT_102824/tree_file")  # Replace "tree_file" with your actual file name.

# Step 1: Check the current taxa names
current_taxa_names <- taxa_names(ps)
# Step 2: Rename taxa names to match ASV format
# Create a new vector for the ASV names (ensure they are unique)
new_taxa_names <- paste0("ASV_", seq_along(current_taxa_names))
# Assign the new names to your new phyloseq object
ps_w_tree<-ps
taxa_names(ps_w_tree) <- new_taxa_names

# Optional: Check the updated taxa names
print(taxa_names(ps_w_tree))

# Check if the tree tips match the new taxa names
tip_labels <- tree$tip.label


# Step 1: Identify taxa names in the phyloseq object that are not in the tree tips
unmatched_taxa <- setdiff(new_taxa_names, tip_labels)

# Step 2: Print unmatched taxa
if (length(unmatched_taxa) > 0) {
    cat("The following taxa do not match any tree tips:\n")
    print(unmatched_taxa)
} else {
    cat("All taxa names match the tree tips.\n")
}



ps_filtered <- subset_taxa(ps_w_tree, taxa_names(ps_w_tree) %in% tip_labels)

# Now, try merging again
physeq_with_tree <- merge_phyloseq(ps_filtered, phy_tree(tree))

```



```{r}

library(vegan)
library(dplyr)
library(ggplot2)
library(phyloseq)
library(tidyverse)
library(dplyr)
library(readxl)

#upload metadata
cognition_metadata <- read_xlsx("/Users/russ/Documents/aim_2_paper_cognition/cognition_metadata.xlsx") #importing the metadata


#creating a 'phyloseq' object
trans_table.no.chimeras<-t(table.no.chimeras)
otu<-otu_table(trans_table.no.chimeras, taxa_are_rows = TRUE)
otu<-t(otu)
samp<-sample_data(cognition_metadata)
row.names(samp) <- samp$seq_file #altering its format so row names are the sample ids
tax_tab<-tax_table(tax)
#phyloseq is essentially a useful bunch of functions for making plots and conducting analyses with this sort of data
ps <- phyloseq(otu, samp, tax_tab) #making the phyloseq object

#removing Chloroplast and mitochondria
ps_no_chloro <- physeq_with_tree %>%
  subset_taxa(
    Kingdom == "Bacteria" &                   #only bacteria
      Family  != "mitochondria" &             #filter out mitochondria
      Class   != "Chloroplast"                #filter out chloroplasts
  )

# Ensure sample IDs match (assuming sample IDs are in the first column of cognition_metadata)
sample_ids <- cognition_metadata[[1]]

# Subset the phyloseq object to keep only the samples that match
ps_cog <- subset_samples(ps_no_chloro, sample_names(ps_no_chloro) %in% sample_ids)

# Check the filtered phyloseq object
ps_cog


#problem solving scores overall  
pretrial_mean<-mean(cognition_metadata$PS_pretrial, na.rm = TRUE)
print(pretrial_mean)

pstake_mean<-mean(cognition_metadata$PS_Test1_solve, na.rm = TRUE)
print(pstake_mean)

ps_post<-mean(cognition_metadata$PS_posttrial_take, na.rm = TRUE)
print(ps_post)

#novelty 
cognition_metadata$NT_test1_take<- suppressWarnings(as.numeric(cognition_metadata$NT_test1_take))

nt_take_mean<-mean(cognition_metadata$NT_test1_take, na.rm = TRUE)
print(nt_take_mean)

#AL
cognition_metadata$meanAL_day2to7_score<- suppressWarnings(as.numeric(cognition_metadata$meanAL_day2to7_score))

al_2_7_mean<-mean(cognition_metadata$meanAL_day2to7_score, na.rm = TRUE)
print(al_2_7_mean)

cognition_metadata$AL_slope<- suppressWarnings(as.numeric(cognition_metadata$AL_slope))

al_slope_mean<-mean(cognition_metadata$AL_slope, na.rm = TRUE)
print(al_slope_mean)

#RL
cognition_metadata$meanRL_day2to7_score<- suppressWarnings(as.numeric(cognition_metadata$meanRL_day2to7_score))

rl_2_7_mean<-mean(cognition_metadata$meanRL_day2to7_score, na.rm = TRUE)
print(rl_2_7_mean)

cognition_metadata$RL_slope<- suppressWarnings(as.numeric(cognition_metadata$RL_slope))

rl_slope_mean<-mean(cognition_metadata$RL_slope, na.rm = TRUE)
print(rl_slope_mean)
```

subsetting the cognition data into species groups subsets 
```{r}
# Filter cognition_metadata for wgs.class 
bcch_data <- cognition_metadata[cognition_metadata$wgs.class == "BCCH", ]
hych_data <- cognition_metadata[cognition_metadata$wgs.class == "HYCH", ]
adv.ca_data <- cognition_metadata[cognition_metadata$wgs.class == "ADV.CA", ]
cach_data <- cognition_metadata[cognition_metadata$wgs.class == "CACH", ]

male_data<-cognition_metadata[cognition_metadata$Sex == "M-DNA", ]
female_data<-cognition_metadata[cognition_metadata$Sex == "F-DNA", ]

#replace dataset with each species group 
# Function to calculate standard error
standard_error <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

# Problem solving scores
pretrial_mean <- mean(female_data$PS_pretrial, na.rm = TRUE)
pretrial_se <- standard_error(female_data$PS_pretrial)
print(paste("Pretrial mean:", pretrial_mean, "SE:", pretrial_se))

pstake_mean <- mean(female_data$PS_Test1_solve, na.rm = TRUE)
pstake_se <- standard_error(female_data$PS_Test1_solve)
print(paste("PStake mean:", pstake_mean, "SE:", pstake_se))

ps_post <- mean(female_data$PS_posttrial_take, na.rm = TRUE)
ps_post_se <- standard_error(female_data$PS_posttrial_take)
print(paste("Posttrial mean:", ps_post, "SE:", ps_post_se))

# Novelty scores
nt_take_mean <- mean(female_data$NT_test1_take, na.rm = TRUE)
nt_take_se <- standard_error(female_data$NT_test1_take)
print(paste("NT Take mean:", nt_take_mean, "SE:", nt_take_se))

# AL scores
al_2_7_mean <- mean(female_data$meanAL_day2to7_score, na.rm = TRUE)
al_2_7_se <- standard_error(female_data$meanAL_day2to7_score)
print(paste("AL 2-7 mean:", al_2_7_mean, "SE:", al_2_7_se))

al_slope_mean <- mean(female_data$AL_slope, na.rm = TRUE)
al_slope_se <- standard_error(female_data$AL_slope)
print(paste("AL Slope mean:", al_slope_mean, "SE:", al_slope_se))

# RL scores
rl_2_7_mean <- mean(female_data$meanRL_day2to7_score, na.rm = TRUE)
rl_2_7_se <- standard_error(female_data$meanRL_day2to7_score)
print(paste("RL 2-7 mean:", rl_2_7_mean, "SE:", rl_2_7_se))

rl_slope_mean <- mean(female_data$RL_slope, na.rm = TRUE)
rl_slope_se <- standard_error(female_data$RL_slope)
print(paste("RL Slope mean:", rl_slope_mean, "SE:", rl_slope_se))


```

creating subsets based on the different tests 
```{r}

#make sure all the values are numeric 
cognition_metadata$PS_posttrial_take<-as.numeric(cognition_metadata$PS_posttrial_take)
cognition_metadata$PS_Test1_solve<-as.numeric(cognition_metadata$PS_Test1_solve)
cognition_metadata$PS_pretrial<-as.numeric(cognition_metadata$PS_pretrial)
cognition_metadata$NT_test1_take<-as.numeric(cognition_metadata$NT_test1_take)
cognition_metadata$meanAL_day2to7_score<-as.numeric(cognition_metadata$meanAL_day2to7_score)
cognition_metadata$meanRL_day2to7_score<-as.numeric(cognition_metadata$meanRL_day2to7_score)
cognition_metadata$AL_slope<-as.numeric(cognition_metadata$AL_slope)
cognition_metadata$RL_slope<-as.numeric(cognition_metadata$RL_slope)



#finding the quantiles of the different groups 
#PROBLEEM SOLVING GROUPS
# Create a basic histogram to see the distribution of PS_Test1_solve
ggplot(cognition_metadata, aes(x = PS_Test1_solve)) +
  geom_histogram(binwidth = 200, fill = "steelblue", color = "black", alpha = 0.7) + # Customize bin width, color, and transparency
  labs(title = "Distribution of PS Test Solve", x = "PS Test Solve", y = "Frequency") + # Axis labels and title
  theme_minimal() # Use a minimal theme for cleaner visualization

# Calculate quantiles
PSquantiles <- quantile(cognition_metadata$PS_Test1_solve, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
PSquantiles

#PSquantiles
#    33%     66% 
# 184.95 1092.50 

# Specify breaks and labels
breaks <- c(-Inf, 185, 1093, 6000)
labels <- c("Good", "Mediocre", "Poor")

# Create 'PS_category' based on 'PS_Test1_solve'
cognition_metadata$PS_category <- cut(cognition_metadata$PS_Test1_solve, breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$PS_category<-cognition_metadata$PS_category

#see total PS_category groups for each species groups
sum_by_category <- cognition_metadata %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(PS_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)



#PROBLEM SOLVING QUANTILES WITHOUT THE 7200s individuals that didn't solve
# Calculate quantiles
PSquantiles <- quantile(filtered_df$PS_Test1_solve, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
PSquantiles

#PSquantiles
#    33%     66% 
# 160.05 505.26 

# Specify breaks and labels
breaks <- c(-Inf, 160.05, 505.26, 7200)
labels <- c("Good", "Mediocre", "Poor")

# Create 'PS_category' based on 'PS_Test1_solve'
filtered_df$PS_category <- cut(filtered_df$PS_Test1_solve, breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$PS_category<-filtered_df$PS_category

#see total PS_category groups for each species groups
sum_by_category <- filtered_df %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(PS_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)









#AL TEST GROUPS
# Create a basic histogram to see the distribution of meanAL_day2to7_score
ggplot(cognition_metadata, aes(x = meanAL_day2to7_score)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "black", alpha = 0.7) + # Customize bin width, color, and transparency
  labs(title = "Distribution of mean AL", x = "AL mean", y = "Frequency") + # Axis labels and title
  theme_minimal() # Use a minimal theme for cleaner visualization

# Calculate quantiles
ALquantiles <- quantile(cognition_metadata$meanAL_day2to7_score, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
ALquantiles

#ALquantiles
#     33%      66% 
# 3.833333 6.833333 

# Specify breaks and labels
breaks <- c(-Inf, 4, 7, 15)
labels <- c("Good", "Mediocre", "Poor")

# Create 'AL_category' based on 'meanAL_day2to7_score'
cognition_metadata$AL_category <- cut(cognition_metadata$meanAL_day2to7_score, breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$AL_category<-cognition_metadata$AL_category

#see total AL_category groups for each species groups
sum_by_category <- cognition_metadata %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(AL_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)



#RL TEST GROUPS
# Create a basic histogram to see the distribution of meanRL_day2to7_score
ggplot(cognition_metadata, aes(x = meanRL_day2to7_score)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "black", alpha = 0.7) + # Customize bin width, color, and transparency
  labs(title = "Distribution of mean RL", x = "RL mean", y = "Frequency") + # Axis labels and title
  theme_minimal() # Use a minimal theme for cleaner visualization

# Calculate quantiles
RLquantiles <- quantile(cognition_metadata$meanRL_day2to7_score, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
RLquantiles

#RLquantiles
#     33%      66% 
# 4.413000 7.043333 

# Specify breaks and labels
breaks <- c(-Inf, 4, 7, 15)
labels <- c("Good", "Mediocre", "Poor")

# Create 'RL_category' based on 'meanRL_day2to7_score'
cognition_metadata$RL_category <- cut(as.numeric(cognition_metadata$meanRL_day2to7_score), breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$RL_category<-cognition_metadata$RL_category

#see total RL_category groups for each species groups
sum_by_category <- cognition_metadata %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(RL_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)


```


finding 0.1% MRA for full phyloseq and each ancestry group
```{r}
library(phyloseq)

ps_cog_rel <- transform_sample_counts(ps_cog, function(x) x / sum(x))

ps_cog_phylum <- tax_glom(ps_cog_rel, taxrank = "Phylum")

library(phyloseq)
library(tidyverse)

# Step 1: Transform to relative abundance
ps_rel <- transform_sample_counts(ps_cog_phylum, function(x) x / sum(x))

# Step 2: Aggregate at Phylum level (skip if already at Phylum level)
ps_phylum <- tax_glom(ps_rel, taxrank = "Phylum")

# Step 3: Convert to long tidy format
df <- psmelt(ps_phylum)

# Step 4: Get phyla with mean relative abundance > 0.001 (i.e., 0.1%) overall
top_phyla_overall <- df %>%
  group_by(Phylum) %>%
  summarise(mean_abundance = mean(Abundance, na.rm = TRUE)) %>%
  filter(mean_abundance > 0.001)

# Step 5: Get phyla with MRA > 0.1% within each wgs.class
top_phyla_by_group <- df %>%
  group_by(wgs.class, Phylum) %>%
  summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  filter(mean_abundance > 0.001)

# Optional: Convert to a list per group if needed
top_phyla_list <- top_phyla_by_group %>%
  group_by(wgs.class) %>%
  summarise(top_phyla = list(unique(Phylum)))



library(tidyverse)

# Replace this with your actual data frame if you're loading from a CSV or other source
MRA_results <- tribble(
  ~wgs.class, ~Phylum, ~mean_abundance,
  "ADV.CA", "Actinobacteriota", 0.026310140,
  "ADV.CA", "Bacteroidota", 0.016451402,
  "ADV.CA", "Firmicutes", 0.546537175,
  "ADV.CA", "Proteobacteria", 0.410322000,
  "BCCH", "Actinobacteriota", 0.029610892,
  "BCCH", "Bacteroidota", 0.026289999,
  "BCCH", "Bdellovibrionota", 0.001371792,
  "BCCH", "Chloroflexi", 0.004957962,
  "BCCH", "Firmicutes", 0.764982376,
  "BCCH", "Planctomycetota", 0.002294112,
  "BCCH", "Proteobacteria", 0.169791862,
  "CACH", "Actinobacteriota", 0.071891454,
  "CACH", "Bacteroidota", 0.015498578,
  "CACH", "Firmicutes", 0.515704365,
  "CACH", "Planctomycetota", 0.002320240,
  "CACH", "Proteobacteria", 0.391341973,
  "CACH", "Verrucomicrobiota", 0.002647933,
  "HYCH", "Actinobacteriota", 0.072525858,
  "HYCH", "Bacteroidota", 0.004154436,
  "HYCH", "Cyanobacteria", 0.005717337,
  "HYCH", "Firmicutes", 0.679628085,
  "HYCH", "Proteobacteria", 0.237332603
)

# Pivot to wide format
wide_MRA_results <- MRA_results %>%
  pivot_wider(names_from = Phylum, values_from = mean_abundance) %>%
  column_to_rownames("wgs.class")  # optional: set rownames

# View the resulting wide-format table
print(wide_MRA_results)

wide_MRA_results<-t(wide_MRA_results)

write.csv(wide_MRA_results, file = "~/Desktop/phyla_abundance_table.csv", row.names = TRUE)



```


PCoA beta Diversity graphs 
```{r}

#choose which distance matrix to use
nmds_cog_jaccard <- ordinate(ps_cog, method = "NMDS", distance = jaccard)
pcoa_cog_jaccard <- ordinate(ps_cog, method = "PCoA", distance = jaccard)

nmds_cog_bray <- ordinate(ps_cog, method = "NMDS", distance = bray)
pcoa_cog_bray <- ordinate(ps_cog, method = "PCoA", distance = bray)

nmds_cog_uniunweight <- ordinate(ps_cog, method = "NMDS", distance = unifrac_unweighted)
pcoa_cog_uniunweight <- ordinate(ps_cog, method = "PCoA", distance = unifrac_unweighted)


# Extract sample data and NMDS scores
nmds_scores <- as.data.frame(nmds_cog_uniunweight$points)
pcoa_scores <- as.data.frame(pcoa_cog_uniunweight$vectors)

sample_data<-ps_cog@sam_data
# Combine sample data and NMDS scores
nmds_data <- cbind(sample_data, nmds_scores)
nmds_data <- subset(nmds_data, 
                    !is.na(wgs.class) & 
                    !is.na(nmds_data$PS_category) & 
                    !is.na(nmds_data$NT_category) & 
                    !is.na(nmds_data$AL_category) & 
                    !is.na(nmds_data$RL_category))


pcoa_data <- cbind(sample_data, pcoa_scores)
pcoa_data <- subset(pcoa_data, 
                    !is.na(wgs.class) & 
                    !is.na(pcoa_data$PS_category) & 
                    !is.na(pcoa_data$NT_category) & 
                    !is.na(pcoa_data$AL_category) & 
                    !is.na(pcoa_data$RL_category))


# Plot NMDS
nmds_plot <- ggplot(nmds_data, aes(x = MDS1, y = MDS2, color = RL_category)) +
  geom_point(size = 3) +
  labs(title = "NMDS of Beta Diversity (UniFrac Unweighted)") +
  theme_minimal() +
  stat_ellipse()


pdf("/Users/russ/Desktop/nmds_plot.pdf", width = 10, height = 10)
print(nmds_plot)
dev.off()

print(nmds_plot)


# PCoA plot
PCoA_plot<-ggplot(pcoa_data, aes(x = Axis.1, y = Axis.2, color = AL_category, shape = meanAL_day2to7_score)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_minimal() +
  labs(
    x = paste0("PCoA Axis 1 (", round(pcoa_data$values$Relative_eig[1] * 100, 1), "% variance)"),
    y = paste0("PCoA Axis 2 (", round(pcoa_data$values$Relative_eig[2] * 100, 1), "% variance)"),
    color = "wgs.class",
    shape = "wgs.class"
  ) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold")
  )+
  stat_ellipse()

pdf("/Users/russ/Desktop/PCoA_plot.pdf", width = 10, height = 10)
print(PCoA_plot)
dev.off()

print(PCoA_plot)


```


DESeq work for bacteria comparisons across cog groups 
```{r}
# Filter out individuals who did not complete the test (PS_Test1_solve == 7200)
ps_cog_filtered <- subset_samples(ps_cog, PS_Test1_solve != 7200)

#Deseq of cog categories (subset the *_category with the test you want to look at)

library(phyloseq)
library(DESeq2)

#replace X_category with whichever test you are looking at results for 

table(sample_data(ps_cog_filtered)$RL_category, useNA = "always")

#make an NA clean phyloseq
ps_cog_clean <- subset_samples(ps_cog_filtered, !is.na(RL_category))

#make sure the levels are correct
sample_data(ps_cog_clean)$RL_category <- factor(sample_data(ps_cog_clean)$RL_category, levels = c("Poor", "Mediocre", "Good"))

# make small numbers and get rid of low abundance bc DESeq2 doesn't handle it well
ps_cog_clean <- filter_taxa(ps_cog_clean, function(x) sum(x > 0) > (0.1 * length(x)), TRUE)

# Create DESeq2 object with RL_category as the design variable
dds <- phyloseq_to_deseq2(ps_cog_clean, ~ RL_category)

# Estimate size factors and dispersions. the "poscounts" This uses positive counts only for estimating size factors, which works much better for sparse datasets like ASV tables. this code skips zeros when estimating size factors, avoiding undefined log operations.
dds <- DESeq(dds, sfType = "poscounts")

summary(dds)

# Good vs Poor
res_good_poor <- results(dds, contrast = c("RL_category", "Good", "Poor"))

# Mediocre vs Poor
res_med_poor <- results(dds, contrast = c("RL_category", "Mediocre", "Poor"))

# Good vs Mediocre
res_good_med <- results(dds, contrast = c("RL_category", "Good", "Mediocre"))

#creates FDR threshold
res_good_poor_sig <- res_good_poor[which(res_good_poor$padj < 0.05), ]


library(ggplot2)

# Extract normalized counts
norm_counts <- counts(dds, normalized = TRUE)
top_taxa <- rownames(res_good_poor_sig)[1:min(10, nrow(res_good_poor_sig))]

# Melt and plot
library(reshape2)
# Step 1: Get normalized counts of top significant ASVs
norm_counts <- counts(dds, normalized = TRUE)
df_DESeq <- as.data.frame(t(norm_counts[top_taxa, ]))

# Step 2: Add SampleID as a column
df_DESeq$SampleID <- rownames(df_DESeq)

# Step 3: Add only the PS_category metadata
meta_df <- data.frame(sample_data(ps_cog_clean))
meta_df$SampleID <- rownames(meta_df)
meta_df <- meta_df[, c("SampleID", "RL_category")]

# Step 4: Merge
df_DESeq <- merge(df_DESeq, meta_df, by = "SampleID")
df_DESeq_melt <- melt(df_DESeq, id.vars = c("SampleID", "RL_category"))

DESeq_RL_good_poor<-ggplot(df_DESeq_melt, aes(x = RL_category, y = value, fill = RL_category)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw() +
  ylab("Normalized Abundance") +
  xlab("RL_category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Save the plot to a PDF
pdf("/Users/russ/Desktop/DESeq_RL_good_poor.pdf", width = 10, height = 10)
print(DESeq_RL_good_poor)
dev.off()


#log transforming for visualization 
log_DESeq_RL<-ggplot(df_DESeq_melt, aes(x = RL_category, y = log10(value + 1), fill = RL_category)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw() +
  ylab("Log10 Normalized Abundance") +
  xlab("RL_category")

# Save the plot to a PDF
pdf("/Users/russ/Desktop/log_DESeq_RL.pdf", width = 10, height = 10)
print(log_DESeq_RL)
dev.off()



#adding the genus name instead of ASV#
# Pull taxonomy table
tax_DESeq <- as.data.frame(tax_table(ps_cog_clean))

# Add a column of ASV IDs to match
tax_DESeq$ASV <- rownames(tax_DESeq)

# For simplicity, keep only ASV and Genus
tax_short <- tax_DESeq[, c("ASV", "Genus")]

# Optional: Fill in missing genus names with ASV IDs
tax_short$Genus[is.na(tax_short$Genus) | tax_short$Genus == ""] <- tax_short$ASV[is.na(tax_short$Genus) | tax_short$Genus == ""]

# Create a lookup vector
asv_to_genus <- setNames(tax_short$Genus, tax_short$ASV)

df_DESeq_melt$Genus <- asv_to_genus[as.character(df_DESeq_melt$variable)]


genus_log_RL<-ggplot(df_DESeq_melt, aes(x = RL_category, y = log10(value + 1), fill = RL_category)) +
  geom_boxplot() +
  facet_wrap(~ Genus, scales = "free_y") +
  theme_bw() +
  ylab("Log10 Normalized Abundance") +
  xlab("RL_category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Save the plot to a PDF
pdf("/Users/russ/Desktop/genus_log_RL.pdf", width = 10, height = 10)
print(genus_log_RL)
dev.off()







#getting the p values 
# convert results to dataframe if needed
res_df <- as.data.frame(res_good_poor)

# Add ASV IDs
res_df$ASV <- rownames(res_df)

# Filter for top_taxa
top_results <- res_df[rownames(res_df) %in% top_taxa, ]

# Add Genus
top_results$Genus <- asv_to_genus[top_results$ASV]

# Select key columns
top_results_clean <- top_results[, c("ASV", "Genus", "log2FoldChange", "pvalue", "padj", "baseMean")]

# Sort by adjusted p-value
top_results_clean <- top_results_clean[order(top_results_clean$padj), ]

print(top_results_clean)

write.csv(top_results_clean, "~/Desktop/top_deseq2_taxa_RL.csv", row.names = FALSE)
```


```{r}
#error bar comparisons 
# Convert results to dataframes and add comparison labels
df_good_poor <- as.data.frame(res_good_poor) %>%
  tibble::rownames_to_column("ASV") %>%
  mutate(Comparison = "Good vs Poor")

df_med_poor <- as.data.frame(res_med_poor) %>%
  tibble::rownames_to_column("ASV") %>%
  mutate(Comparison = "Mediocre vs Poor")

df_good_med <- as.data.frame(res_good_med) %>%
  tibble::rownames_to_column("ASV") %>%
  mutate(Comparison = "Good vs Mediocre")

# Combine
df_all <- bind_rows(df_good_poor, df_med_poor, df_good_med)

tax_DESeq <- as.data.frame(tax_table(ps_cog_clean)) %>%
  tibble::rownames_to_column("ASV")

df_all <- left_join(df_all, tax_DESeq[, c("ASV", "Genus")], by = "ASV")

# Replace missing Genus names with ASV IDs
df_all$Genus[is.na(df_all$Genus) | df_all$Genus == ""] <- df_all$ASV[is.na(df_all$Genus) | df_all$Genus == ""]


library(ggplot2)
#fix this and fill in with the signifianct ASVs found from the DESeq2 run top_results_clean
df_sig_asvs <- df_all %>% filter(ASV %in% c("ASV_47", "ASV_119", "ASV_548"))


RL_sig_ASVs<-ggplot(df_sig_asvs, aes(x = log2FoldChange, y = Genus)) +
  geom_point(color = "black") +
  geom_errorbarh(aes(xmin = log2FoldChange - lfcSE,
                     xmax = log2FoldChange + lfcSE),
                 height = 0.2, color = "black") +
  facet_wrap(~ Comparison, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw() +
  xlab("Log2 Fold Change ± s.e.") +
  ylab("")

# Save the plot to a PDF
pdf("/Users/russ/Desktop/RL_sig_ASVs.pdf", width = 10, height = 5)
print(RL_sig_ASVs)
dev.off()


```



#POST HOC STUFF AT ASV LEVEL
post hoc for the comparison of sig ASVs between cognition groups 
```{r}

# Identify significant ASVs from each comparison
sig_good_poor <- rownames(res_good_poor)[which(res_good_poor$padj < 0.05)]
sig_med_poor  <- rownames(res_med_poor)[which(res_med_poor$padj < 0.05)]
sig_good_med  <- rownames(res_good_med)[which(res_good_med$padj < 0.05)]

# Union of all significant ASVs
sig_asvs <- unique(c(sig_good_poor, sig_med_poor, sig_good_med))


# Create data frame with post hoc comparison results
posthoc_df <- data.frame(
  ASV = sig_asvs,
  log2FC_good_poor = res_good_poor[sig_asvs, "log2FoldChange"],
  padj_good_poor = res_good_poor[sig_asvs, "padj"],
  
  log2FC_med_poor = res_med_poor[sig_asvs, "log2FoldChange"],
  padj_med_poor = res_med_poor[sig_asvs, "padj"],
  
  log2FC_good_med = res_good_med[sig_asvs, "log2FoldChange"],
  padj_good_med = res_good_med[sig_asvs, "padj"]
)


# Add Genus column
posthoc_df$Genus <- tax_short$Genus[match(posthoc_df$ASV, tax_short$ASV)]

view(posthoc_df)

```









ancestry DESeq2 

```{r}
library(phyloseq)
library(DESeq2)
library(ggplot2)
library(reshape2)

# Filter out individuals who did not complete the test
ps_anc_filtered <- subset_samples(ps_cog, PS_Test1_solve != 7200)

# Remove NAs in ancestry grouping
ps_anc_clean <- subset_samples(ps_anc_filtered, !is.na(wgs.class))

# Ensure ancestry group levels are correctly set
sample_data(ps_anc_clean)$wgs.class <- factor(sample_data(ps_anc_clean)$wgs.class)

# Filter low-abundance ASVs
ps_anc_clean <- filter_taxa(ps_anc_clean, function(x) sum(x > 0) > (0.1 * length(x)), TRUE)

# Create DESeq2 object with ancestry group as design
dds_anc <- phyloseq_to_deseq2(ps_anc_clean, ~ wgs.class)

# DESeq with poscounts normalization
dds_anc <- DESeq(dds_anc, sfType = "poscounts")

res_adv_bcch <- results(dds_anc, contrast = c("wgs.class", "ADV.CA", "BCCH"))
res_cach_hych <- results(dds_anc, contrast = c("wgs.class", "CACH", "HYCH"))
res_adv_hych <- results(dds_anc, contrast = c("wgs.class", "ADV.CA", "HYCH"))
res_bcch_cach <- results(dds_anc, contrast = c("wgs.class", "BCCH", "CACH"))
res_bcch_hych <- results(dds_anc, contrast = c("wgs.class", "BCCH", "HYCH"))
res_cach_adv <- results(dds_anc, contrast = c("wgs.class", "CACH", "ADV.CA"))

# Get significant ASVs for one contrast (e.g., ADV.CA vs BCCH)
res_sig <- res_adv_bcch[which(res_adv_bcch$padj < 0.05), ]

# Extract normalized counts
norm_counts <- counts(dds_anc, normalized = TRUE)
top_taxa <- rownames(res_sig)[1:min(10, nrow(res_sig))]

# Prepare plotting data
df_DESeq <- as.data.frame(t(norm_counts[top_taxa, ]))
df_DESeq$SampleID <- rownames(df_DESeq)

meta_df <- data.frame(sample_data(ps_anc_clean))
meta_df$SampleID <- rownames(meta_df)
meta_df <- meta_df[, c("SampleID", "wgs.class")]

df_DESeq <- merge(df_DESeq, meta_df, by = "SampleID")
df_DESeq_melt <- melt(df_DESeq, id.vars = c("SampleID", "wgs.class"))

# Plot raw and log-transformed counts
ggplot(df_DESeq_melt, aes(x = wgs.class, y = value, fill = wgs.class)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw() +
  ylab("Normalized Abundance") +
  xlab("Ancestry Group")

ggplot(df_DESeq_melt, aes(x = wgs.class, y = log10(value + 1), fill = wgs.class)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw() +
  ylab("Log10 Normalized Abundance") +
  xlab("Ancestry Group")

# Add genus labels
tax_df <- as.data.frame(tax_table(ps_anc_clean))
tax_df$ASV <- rownames(tax_df)
tax_short <- tax_df[, c("ASV", "Genus")]
tax_short$Genus[is.na(tax_short$Genus) | tax_short$Genus == ""] <- tax_short$ASV[is.na(tax_short$Genus) | tax_short$Genus == ""]
asv_to_genus <- setNames(tax_short$Genus, tax_short$ASV)
df_DESeq_melt$Genus <- asv_to_genus[as.character(df_DESeq_melt$variable)]

# Log-scale genus-level boxplots
ggplot(df_DESeq_melt, aes(x = wgs.class, y = log10(value + 1), fill = wgs.class)) +
  geom_boxplot() +
  facet_wrap(~ Genus, scales = "free_y") +
  theme_bw() +
  ylab("Log10 Normalized Abundance") +
  xlab("Ancestry Group")

# Export top taxa table
res_df <- as.data.frame(res_adv_bcch)
res_df$ASV <- rownames(res_df)
top_results <- res_df[rownames(res_df) %in% top_taxa, ]
top_results$Genus <- asv_to_genus[top_results$ASV]
top_results_clean <- top_results[, c("ASV", "Genus", "log2FoldChange", "pvalue", "padj", "baseMean")]
top_results_clean <- top_results_clean[order(top_results_clean$padj), ]
write.csv(top_results_clean, "~/Desktop/top_deseq2_taxa_wgsclass.csv", row.names = FALSE)

```

post hoc of ancestry ASVs
```{r}

# Your original comparisons
res_adv_bcch <- results(dds_anc, contrast = c("wgs.class", "ADV.CA", "BCCH"))
res_cach_hych <- results(dds_anc, contrast = c("wgs.class", "CACH", "HYCH"))
res_adv_hych <- results(dds_anc, contrast = c("wgs.class", "ADV.CA", "HYCH"))

res_bcch_cach <- results(dds_anc, contrast = c("wgs.class", "BCCH", "CACH"))
res_bcch_hych <- results(dds_anc, contrast = c("wgs.class", "BCCH", "HYCH"))
res_cach_adv <- results(dds_anc, contrast = c("wgs.class", "CACH", "ADV.CA"))

# Identify significant ASVs (padj < 0.05) from each comparison
sig_adv_bcch <- rownames(res_adv_bcch)[which(res_adv_bcch$padj < 0.05)]
sig_cach_hych <- rownames(res_cach_hych)[which(res_cach_hych$padj < 0.05)]
sig_adv_hych  <- rownames(res_adv_hych)[which(res_adv_hych$padj < 0.05)]

sig_bcch_cach <- rownames(res_bcch_cach)[which(res_bcch_cach$padj < 0.05)]
sig_bcch_hych <- rownames(res_bcch_hych)[which(res_bcch_hych$padj < 0.05)]
sig_cach_adv  <- rownames(res_cach_adv)[which(res_cach_adv$padj < 0.05)]

# Combine all significant ASVs across all comparisons
sig_asvs_ancestry <- unique(c(
  sig_adv_bcch, sig_cach_hych, sig_adv_hych,
  sig_bcch_cach, sig_bcch_hych, sig_cach_adv
))

# Create a dataframe with log2 fold changes and adjusted p-values for all comparisons
posthoc_ancestry_df <- data.frame(
  ASV = sig_asvs_ancestry,
  
  log2FC_adv_bcch = res_adv_bcch[sig_asvs_ancestry, "log2FoldChange"],
  padj_adv_bcch   = res_adv_bcch[sig_asvs_ancestry, "padj"],
  
  log2FC_cach_hych = res_cach_hych[sig_asvs_ancestry, "log2FoldChange"],
  padj_cach_hych   = res_cach_hych[sig_asvs_ancestry, "padj"],
  
  log2FC_adv_hych = res_adv_hych[sig_asvs_ancestry, "log2FoldChange"],
  padj_adv_hych   = res_adv_hych[sig_asvs_ancestry, "padj"],
  
  log2FC_bcch_cach = res_bcch_cach[sig_asvs_ancestry, "log2FoldChange"],
  padj_bcch_cach   = res_bcch_cach[sig_asvs_ancestry, "padj"],
  
  log2FC_bcch_hych = res_bcch_hych[sig_asvs_ancestry, "log2FoldChange"],
  padj_bcch_hych   = res_bcch_hych[sig_asvs_ancestry, "padj"],
  
  log2FC_cach_adv = res_cach_adv[sig_asvs_ancestry, "log2FoldChange"],
  padj_cach_adv   = res_cach_adv[sig_asvs_ancestry, "padj"]
)

# Add Genus column from taxonomic table
posthoc_ancestry_df$Genus <- tax_short$Genus[match(posthoc_ancestry_df$ASV, tax_short$ASV)]

# View the final post hoc results
View(posthoc_ancestry_df)



```



subsetting the alpha diversity measures into it's own dataset 
```{r}
library(vegan)
library(dplyr)
library(ggplot2)
library(phyloseq)
library(tidyverse)

# Bray-Curtis dissimilarity
bray <- phyloseq::distance(ps_cog, method = "bray")

# Jaccard dissimilarity
jaccard <- phyloseq::distance(ps_cog, method = "jaccard")

# Unweighted UniFrac
unifrac_unweighted <- phyloseq::distance(ps_cog, method = "unifrac")

# Weighted UniFrac
unifrac_weighted <- phyloseq::distance(ps_cog, method = "wunifrac")


#adding faith's phylo diversity 
# Load required library
library(picante)
library(ape)

# Root the tree
phy_tree_cog <- phy_tree(ps_cog)
rooted_tree <- root(phy_tree_cog, outgroup = phy_tree_cog$tip.label[1], resolve.root = TRUE)

# Replace the tree in the phyloseq object (optional, if you want to work with the rooted tree in ps_cog)
phy_tree(ps_cog) <- rooted_tree


# Extract tree and OTU table from ps_cog
phy_tree_cog <- phy_tree(ps_cog)
otu_cog <- as.matrix(otu_table(ps_cog))

# Calculate Faith's PD
faith_pd <- pd(otu_cog, phy_tree_cog)$PD

# Create the diversity dataframe
diversity_df <- data.frame(
  SampleID = sample_names(ps_cog),
  Shannon = estimate_richness(ps_cog, measures = "Shannon")[, "Shannon"],
  Chao = estimate_richness(ps_cog, measures = "Chao1")[, "Chao1"],
  InvSimpson = estimate_richness(ps_cog, measures = "InvSimpson")[, "InvSimpson"],
  Richness = estimate_richness(ps_cog, measures = "Observed")[, "Observed"],
  OTUs = rowSums(otu_table(ps_cog)),
  FaithPD = faith_pd
)


#making the metadata object and confirming dataframes 
meta_data<-sample_data(ps_cog)
meta_data <- data.frame(as(meta_data, "matrix"))
diversity_df<-as.data.frame(diversity_df)

# Rename seq_file to SampleID in meta_data
colnames(meta_data)[colnames(meta_data) == "seq_file"] <- "SampleID"

# Combine with metadata
diversity_df <- merge(diversity_df, meta_data, by = "SampleID")


#SEX DIFFERENCES filtering 
# Filter for all males
diversity_df_male <- diversity_df %>%
  filter(Sex == "M-DNA")
# Filter for all females
diversity_df_female <- diversity_df %>%
  filter(Sex == "F-DNA")

#sex subsetting phyloseq 
ps_cog_male <- subset_samples(ps_cog, Sex == "M-DNA")
ps_cog_female <- subset_samples(ps_cog, Sex == "F-DNA")

# Bray-Curtis dissimilarity
bray_female <- phyloseq::distance(ps_cog_female, method = "bray")
bray_male <- phyloseq::distance(ps_cog_male, method = "bray")
# Jaccard dissimilarity
jaccard_female <- phyloseq::distance(ps_cog_female, method = "jaccard")
jaccard_male <- phyloseq::distance(ps_cog_male, method = "jaccard")
# Unweighted UniFrac
unifrac_unweighted_male <- phyloseq::distance(ps_cog_male, method = "unifrac")
unifrac_unweighted_female <- phyloseq::distance(ps_cog_female, method = "unifrac")
# Weighted UniFrac
unifrac_weighted_male <- phyloseq::distance(ps_cog_male, method = "wunifrac")
unifrac_weighted_female <- phyloseq::distance(ps_cog_female, method = "wunifrac")

```


stats
```{r}
#checking for normal distribution
hist(diversity_df$Shannon, main="Histogram of Shannon Diversity", xlab="Shannon Diversity", col="lightblue")
shapiro.test(diversity_df$Shannon)

#making a no NA dataset 
df_clean <- diversity_df[!is.na(diversity_df$PS_Test1_solve), ]

library(MASS)
#checking the best fit for the family funciton of the GLMM 
fit1<-fitdistr(df_clean$PS_Test1_solve[-170], "normal")
fit2<-fitdistr(df_clean$PS_Test1_solve[-170], "lognormal")
fit3<-fitdistr(df_clean$PS_Test1_solve[-170], "exponential")
fit4<-fitdistr(df_clean$PS_Test1_solve[-170], "poisson")


AIC(fit1,fit2,fit3,fit4)


#for PS_Test1_solve the best fit was lognormal
#fit1	2	1217.935		
#fit2	2	1088.745		
#fit3	1	1119.319		
#fit4	1	185343.953	

#for NT the best fit was also lognormal 
#fit1	2	945.9896		
#fit2	2	922.0772		
#fit3	1	936.5560		
#fit4	1	31014.7289	

#for AL it was lognormal
#fit1	2	337.8788		
#fit2	2	335.7160		
#fit3	1	361.7500		
#fit4	1	Inf	

# for RL it was barely lognormal 
#fit1	2	285.8065		
#fit2	2	284.1997		
#fit3	1	343.8364		
#fit4	1	Inf	

```


glmer for ALPHA testing
```{r}
library(lme4)
library(car)

diversity_df$PS_posttrial_take<-as.numeric(diversity_df$PS_posttrial_take)
filtered_df$PS_Test1_solve<-as.numeric(filtered_df$PS_Test1_solve)
diversity_df$PS_pretrial<-as.numeric(diversity_df$PS_pretrial)
diversity_df$meanAL_day2to7_score<-as.numeric(diversity_df$meanAL_day2to7_score)
diversity_df$meanRL_day2to7_score<-as.numeric(diversity_df$meanRL_day2to7_score)

log_diversity_df<-df_clean

#log transforming all based on the best fit model 
#log_diversity_df$log_PS_Test1_solve <- log(df_clean$PS_Test1_solve)
#log_diversity_df$log_meanAL_day2to7_score <- log(df_clean$meanAL_day2to7_score)
#log_diversity_df$log_meanRL_day2to7_score <- log(df_clean$meanRL_day2to7_score)


#SHANNON DIVERSITY
#simplified version
shan.cog <- lmer(PS_Test1_solve ~ Shannon*wgs.class*Sex  + (1|Nest/Collection_year) + (1|Nest_Site), data = filtered_df)
summary(shan.cog)
#check the Anova 
Anova(shan.cog)

# First round of model simplification
shan.cog.new1 <- update(shan.cog, ~ . 
  - wgs.class:Sex
  - Shannon:Sex
  - Shannon:wgs.class
  -Shannon:wgs.class:Sex)

summary(shan.cog.new1)
anova(shan.cog,shan.cog.new1)
Anova(shan.cog.new1)

#CHAO DIVERSITY
#simplified version
chao.cog <- lmer(PS_Test1_solve ~ Chao*wgs.class*Sex + (1|Nest/Collection_year) + (1|Nest_Site), data = filtered_df)
summary(chao.cog)
#check the Anova 
Anova(chao.cog)

# First round of model simplification
chao.cog.new1 <- update(chao.cog, ~ . 
  - wgs.class:Sex
  - wgs.class:Chao
  - Chao:Sex
  - Chao:Sex:wgs.class)

summary(chao.cog.new1)
anova(chao.cog,chao.cog.new1)
Anova(chao.cog.new1)



#NUMBER OF ASVs
#simplified version
asvnum.cog <- lmer(meanRL_day2to7_score ~ OTUs*wgs.class*Sex  + (1|Nest/Collection_year) + (1|Nest_Site), data = diversity_df)
summary(asvnum.cog)
#check the Anova 
Anova(asvnum.cog)

# First round of model simplification
asvnum.cog.new1 <- update(asvnum.cog, ~ . 
  - OTUs:Sex
  - wgs.class:Sex
  - OTUs:wgs.class:Sex
  - OTUs:wgs.class)

summary(asvnum.cog.new1)
anova(asvnum.cog,asvnum.cog.new1)
Anova(asvnum.cog.new1)



#Faith's phylo diversity
#simplified version
faith.cog <- lmer(meanRL_day2to7_score ~ FaithPD*wgs.class*Sex  + (1|Nest/Collection_year) + (1|Nest_Site), data = diversity_df)
summary(faith.cog)
#check the Anova 
Anova(faith.cog)

# First round of model simplification
faith.cog.new1 <- update(faith.cog, ~ . 
  -FaithPD:Sex
  -FaithPD:wgs.class:Sex
  -wgs.class:Sex
  -FaithPD:wgs.class)

summary(faith.cog.new1)
anova(faith.cog,faith.cog.new1)
Anova(faith.cog.new1)


#####



```


PERMANOVA for BETA diversity testing 
```{r}
#perform PERMANOVA
#testing bray jaccard unifrac_weighted unifrac_unweighted
perm_cog_df <- data.frame(sample_data(ps_cog))

permanova_bray <- adonis2(bray ~ PS_category+Sex+wgs.class , 
                            data = perm_cog_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_bray)



# Impute missing values in PS_Test1_solve with the mean
diversity_clean<-diversity_df

diversity_clean$PS_Test1_solve[is.na(diversity_clean$PS_Test1_solve)] <- 
  mean(diversity_clean$PS_Test1_solve, na.rm = TRUE)
# Impute missing values in Sex with the most frequent category
diversity_clean$Sex[is.na(diversity_clean$Sex)] <- 
  names(sort(table(diversity_clean$Sex), decreasing = TRUE))[1]
# Impute missing values in wgs.class with the most frequent category
diversity_clean$wgs.class[is.na(diversity_clean$wgs.class)] <- 
  names(sort(table(diversity_clean$wgs.class), decreasing = TRUE))[1]

# Perform db-RDA
db_rda <- vegan::capscale(unifrac_weighted ~ PS_Test1_solve + Sex + wgs.class, 
                           data = diversity_clean, 
                           distance = "bray")
# Summarize results
summary(db_rda)
# ANOVA-like test to assess significance
anova(db_rda, permutations = 10000)





permanova_jaccard <- adonis2(jaccard ~ meanRL_day2to7_score + Sex + wgs.class, 
                            data = diversity_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_jaccard)


permanova_unifrac_unweighted <- adonis2(unifrac_unweighted ~ meanRL_day2to7_score + Sex + wgs.class, 
                            data = diversity_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_unifrac_unweighted)

permanova_unifrac_weighted <- adonis2(unifrac_weighted ~ meanRL_day2to7_score + Sex + wgs.class, 
                            data = diversity_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_unifrac_weighted)
```



graphs of significant interactions and aspects of alpha diversity
```{r}
library(cowplot)

#filter out NAs in wgs.class column 
library(dplyr)
PS_Faith_graph_df <- filtered_df %>%
  filter(!is.na(wgs.class))

#log PS and faith with ancestry
PS_faith_wgs_sex_ <- ggplot(PS_Faith_graph_df, aes(x = FaithPD, y = PS_Test1_solve, color = wgs.class)) +
  geom_point(size = 3) +
  geom_smooth(
    method = "lm",
    se = FALSE,  # Add 95% confidence intervals
    aes(linetype = wgs.class),  # Maintain separate lines and CIs by sex
    size = 1
  ) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash")) +  # Adjust for number of groups
  labs(
    x = "Faith's Phylogenetic Diversity", 
    y = "Problem Solving Score", 
    color = "Ancestry Group", 
    linetype = "Ancestry Group"
  ) +
  theme_minimal() +
  theme_cowplot(12)


pdf("/Users/russ/Desktop/PS_shan_wgs_sex_.pdf", width = 7, height = 5)
print(PS_faith_wgs_sex_)
dev.off()


#same problem solving graph as above but no individuals that did not complete the test (scores of 7200 seconds)
# Filter out individuals with a score of 7200
filtered_df <- diversity_df %>% 
  filter(PS_Test1_solve != 7200)

# Then use the filtered data in your plot
PS_faith_wgs_sex_filt <- ggplot(filtered_df, aes(x = FaithPD, y = PS_Test1_solve, color = wgs.class)) +
  geom_point(size = 3) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    aes(linetype = wgs.class),
    size = 1
  ) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash")) +
  labs(
    x = "Faith's Phylogenetic Diversity", 
    y = "Problem Solving Score", 
    color = "Ancestry Group", 
    linetype = "Ancestry Group"
  ) +
  theme_minimal() +
  theme_cowplot(12)


print(PS_faith_wgs_sex_filt)

#slopes of the lines in the graph 
library(emmeans)
model <- lm(PS_Test1_solve ~ FaithPD * wgs.class, data = PS_Faith_graph_df)

emmeans_results <- emtrends(model, spec = "wgs.class", var = "FaithPD")
view(emmeans_results)

```


ASV and RL test comparison 
```{r}
otu_wgs_sex <- ggplot(diversity_df, aes(x = OTUs, y = meanRL_day2to7_score, color = Sex, shape = Sex)) +
  geom_point(size = 3) +
  geom_smooth(
    method = "lm", 
    se = TRUE,  # Show 95% confidence intervals
    aes(linetype = Sex), 
    size = 1
  ) + 
  scale_linetype_manual(values = c("solid", "dashed")) +  # Assuming two sexes
  labs(
    x = "Number of ASVs", 
    y = "Reversal Learning Score", 
    color = "Sex", 
    shape = "Sex", 
    linetype = "Sex"
  ) +
  theme_minimal() +
  theme_cowplot(12) +
  facet_wrap(~wgs.class)



pdf("/Users/russ/Desktop/otu_wgs_sex.pdf", width = 10, height = 10)
print(otu_wgs_sex)
dev.off()



#filter out NAs in Sex column 
library(dplyr)
ASV_graph_df <- diversity_df %>%
  filter(!is.na(Sex))

#same comparison as above with OTUs but sex panels and wgs lines instead 
wgs_by_sex<-ggplot(ASV_graph_df, aes(x = OTUs, y = meanRL_day2to7_score, color = wgs.class, linetype = wgs.class)) +
  geom_point(aes(shape = wgs.class), size = 3) +
  geom_smooth(
    method = "lm", 
    se = FALSE,  # <-- removes shaded confidence intervals
    size = 1
  ) + 
  scale_linetype_manual(values = c("solid", "dashed", "dotdash", "twodash")) +
  labs(
    x = "Number of ASVs", 
    y = "Reversal Learning Score", 
    color = "Ancestry Group", 
    shape = "Ancestry Group", 
    linetype = "Ancestry Group"
  ) +
  theme_minimal() +
  theme_cowplot(12) +
  facet_wrap(~Sex)

pdf("/Users/russ/Desktop/wgs_by_sex.pdf", width = 10, height = 5)
print(wgs_by_sex)
dev.off()




#just the ASVs and the reversal learning significant relationship graphed
ASV_RL_graph<-ggplot(diversity_df, aes(x = OTUs, y = meanRL_day2to7_score)) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue", size = 1) +
  labs(
    x = "Number of ASVs", 
    y = "Reversal Learning Score"
  ) +
  theme_minimal() +
  theme_cowplot(12)

pdf("/Users/russ/Desktop/ASV_RL_graph.pdf", width = 10, height = 5)
print(ASV_RL_graph)
dev.off()

```


```{r}
faith_RL <- ggplot(log_diversity_df, aes(x = FaithPD, y = log_meanRL_day2to7_score)) +
  geom_point(aes(size = 3)) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    size = 1
  ) +
  labs(
    x = "Faith's Phylogenetic Diversity",
    y = "Reversal Learning Score",
    shape = "Sex"
  ) +
  theme_minimal() +
  theme_cowplot(12)



pdf("/Users/russ/Desktop/faith_RL.pdf", width = 10, height = 10)
print(faith_RL)
dev.off()


#bootstrapping the outliers for the graph above and regraphing 
# Filter out rows with NA in relevant variables
log_diversity_clean <- log_diversity_df %>%
  filter(!is.na(FaithPD) & !is.na(log_meanRL_day2to7_score))

# Fit the model on cleaned data
fit <- lm(log_meanRL_day2to7_score ~ FaithPD, data = log_diversity_clean)

# Calculate Cook's distance - bootstrapping using Cook's 
cooksd <- cooks.distance(fit)

# Define threshold
outlier_threshold <- 4 / nrow(log_diversity_clean)

# Flag outliers
log_diversity_clean$outlier <- cooksd > outlier_threshold




library(ggplot2)
library(dplyr)
library(cowplot)
library(boot)

# Keep only inliers for bootstrapping
inliers <- log_diversity_clean %>% filter(!outlier)

# Define bootstrap function
boot_lm <- function(data, indices) {
  d <- data[indices, ]
  fit <- lm(log_meanRL_day2to7_score ~ FaithPD, data = d)
  coef(fit)
}

# Run bootstrap (e.g., 1000 iterations)
set.seed(123)
boot_results <- boot(data = inliers, statistic = boot_lm, R = 1000)

# Generate bootstrapped prediction lines
faith_seq <- seq(min(inliers$FaithPD), max(inliers$FaithPD), length.out = 100)

pred_lines <- lapply(1:nrow(boot_results$t), function(i) {
  intercept <- boot_results$t[i, 1]
  slope <- boot_results$t[i, 2]
  data.frame(
    FaithPD = faith_seq,
    Predicted = intercept + slope * faith_seq
  )
})



library(ggplot2)
library(cowplot)

# Base plot with outliers highlighted
faith_RL <- ggplot(log_diversity_clean, aes(x = FaithPD, y = log_meanRL_day2to7_score)) +
  geom_point(aes(shape = Sex, color = outlier), size = 3) +
  scale_color_manual(values = c("black", "red")) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black",
    size = 1
  ) +
  labs(
    x = "Faith's Phylogenetic Diversity",
    y = "Reversal Learning Score",
    shape = "Sex",
    color = "Outlier"
  ) +
  theme_minimal() +
  theme_cowplot(12)

# Add bootstrapped lines
for (line in pred_lines) {
  faith_RL <- faith_RL + 
    geom_line(data = line, aes(x = FaithPD, y = Predicted),
              inherit.aes = FALSE,
              alpha = 0.05,
              color = "gray")
}

# Show the plot
faith_RL







```


post hoc work 
```{r}
library(emmeans)

# Get pairwise comparisons for the interaction
model <- lm(PS_Test1_solve ~ FaithPD * wgs.class, data = filtered_df)

emmeans_results <- emtrends(model, spec = "wgs.class", var = "FaithPD")
view(emmeans_results)

emmeans_results <- emtrends(shan.cog, specs = ~ wgs.class * Sex, var = "Shannon")




#post hoc comparison for ASVs in the RL test keeping contrasts within sex for ancestry differences 

model <- lmer(log(meanRL_day2to7_score)~OTUs*wgs.class*Sex + (1|Collection_year/Nest) + (1|Nest), data=diversity_df)
ABC <- emtrends(model, pairwise ~ wgs.class:Sex, var= "OTUs", by= "Sex")
summary(ABC)




#comparing the slopes 
# Fit the interaction model
model <- lm(PS_Test1_solve ~ FaithPD * wgs.class, data = filtered_df)
summary(model)

library(emmeans)

# Estimate the FaithPD slope per group
emtrends_model <- emtrends(model, var = "FaithPD", specs = "wgs.class")

# Pairwise comparisons of those slopes
pairs(emtrends_model)




```

tables for the birds that completed vs did not complete the behavioral trials 
```{r}

table(ps_cog@sam_data$PS_Test1_solve)

```





