---
title: "Full_Cognition_Script"
author: "Austin Russell"
date: "2025-04-23"
output: html_document
---

```{r}

#creating a 'phyloseq' object
trans_table.no.chimeras<-t(table.no.chimeras)
otu<-otu_table(trans_table.no.chimeras, taxa_are_rows = TRUE)
otu<-t(otu)
samp<-sample_data(metadata)
row.names(samp) <- samp$seq_file #altering its format so row names are the sample ids
tax_tab<-tax_table(tax)
#phyloseq is essentially a useful bunch of functions for making plots and conducting analyses with this sort of data
ps <- phyloseq(otu, samp, tax_tab) #making the phyloseq object

#removing Chloroplast and mitochondria
ps_no_chloro <- physeq_with_tree %>%
  subset_taxa(
    Kingdom == "Bacteria" &                   #only bacteria
      Family  != "mitochondria" &             #filter out mitochondria
      Class   != "Chloroplast"                #filter out chloroplasts
  )

# Ensure sample IDs match (assuming sample IDs are in the first column of cognition_metadata)
sample_ids <- cognition_metadata[[1]]

# Subset the phyloseq object to keep only the samples that match
ps_cog <- subset_samples(ps_no_chloro, sample_names(ps_no_chloro) %in% sample_ids)

# Check the filtered phyloseq object
ps_cog


#problem solving scores overall  
pretrial_mean<-mean(cognition_metadata$PS_pretrial, na.rm = TRUE)
print(pretrial_mean)

pstake_mean<-mean(cognition_metadata$PS_Test1_solve, na.rm = TRUE)
print(pstake_mean)

ps_post<-mean(cognition_metadata$PS_posttrial_take, na.rm = TRUE)
print(ps_post)

#novelty 
cognition_metadata$NT_test1_take<- suppressWarnings(as.numeric(cognition_metadata$NT_test1_take))

nt_take_mean<-mean(cognition_metadata$NT_test1_take, na.rm = TRUE)
print(nt_take_mean)

#AL
cognition_metadata$meanAL_day2to7_score<- suppressWarnings(as.numeric(cognition_metadata$meanAL_day2to7_score))

al_2_7_mean<-mean(cognition_metadata$meanAL_day2to7_score, na.rm = TRUE)
print(al_2_7_mean)

cognition_metadata$AL_slope<- suppressWarnings(as.numeric(cognition_metadata$AL_slope))

al_slope_mean<-mean(cognition_metadata$AL_slope, na.rm = TRUE)
print(al_slope_mean)

#RL
cognition_metadata$meanRL_day2to7_score<- suppressWarnings(as.numeric(cognition_metadata$meanRL_day2to7_score))

rl_2_7_mean<-mean(cognition_metadata$meanRL_day2to7_score, na.rm = TRUE)
print(rl_2_7_mean)

cognition_metadata$RL_slope<- suppressWarnings(as.numeric(cognition_metadata$RL_slope))

rl_slope_mean<-mean(cognition_metadata$RL_slope, na.rm = TRUE)
print(rl_slope_mean)
```

subsetting the cognition data into species groups subsets 
```{r}
# Filter cognition_metadata for wgs.class 
bcch_data <- cognition_metadata[cognition_metadata$wgs.class == "BCCH", ]
hych_data <- cognition_metadata[cognition_metadata$wgs.class == "HYCH", ]
adv.ca_data <- cognition_metadata[cognition_metadata$wgs.class == "ADV.CA", ]
cach_data <- cognition_metadata[cognition_metadata$wgs.class == "CACH", ]

male_data<-cognition_metadata[cognition_metadata$Sex == "M-DNA", ]
female_data<-cognition_metadata[cognition_metadata$Sex == "F-DNA", ]

#replace dataset with each species group 
# Function to calculate standard error
standard_error <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

# Problem solving scores
pretrial_mean <- mean(female_data$PS_pretrial, na.rm = TRUE)
pretrial_se <- standard_error(female_data$PS_pretrial)
print(paste("Pretrial mean:", pretrial_mean, "SE:", pretrial_se))

pstake_mean <- mean(female_data$PS_Test1_solve, na.rm = TRUE)
pstake_se <- standard_error(female_data$PS_Test1_solve)
print(paste("PStake mean:", pstake_mean, "SE:", pstake_se))

ps_post <- mean(female_data$PS_posttrial_take, na.rm = TRUE)
ps_post_se <- standard_error(female_data$PS_posttrial_take)
print(paste("Posttrial mean:", ps_post, "SE:", ps_post_se))

# Novelty scores
nt_take_mean <- mean(female_data$NT_test1_take, na.rm = TRUE)
nt_take_se <- standard_error(female_data$NT_test1_take)
print(paste("NT Take mean:", nt_take_mean, "SE:", nt_take_se))

# AL scores
al_2_7_mean <- mean(female_data$meanAL_day2to7_score, na.rm = TRUE)
al_2_7_se <- standard_error(female_data$meanAL_day2to7_score)
print(paste("AL 2-7 mean:", al_2_7_mean, "SE:", al_2_7_se))

al_slope_mean <- mean(female_data$AL_slope, na.rm = TRUE)
al_slope_se <- standard_error(female_data$AL_slope)
print(paste("AL Slope mean:", al_slope_mean, "SE:", al_slope_se))

# RL scores
rl_2_7_mean <- mean(female_data$meanRL_day2to7_score, na.rm = TRUE)
rl_2_7_se <- standard_error(female_data$meanRL_day2to7_score)
print(paste("RL 2-7 mean:", rl_2_7_mean, "SE:", rl_2_7_se))

rl_slope_mean <- mean(female_data$RL_slope, na.rm = TRUE)
rl_slope_se <- standard_error(female_data$RL_slope)
print(paste("RL Slope mean:", rl_slope_mean, "SE:", rl_slope_se))


```

creating subsets based on the different tests 
```{r}

#make sure all the values are numeric 
cognition_metadata$PS_posttrial_take<-as.numeric(cognition_metadata$PS_posttrial_take)
cognition_metadata$PS_Test1_solve<-as.numeric(cognition_metadata$PS_Test1_solve)
cognition_metadata$PS_pretrial<-as.numeric(cognition_metadata$PS_pretrial)
cognition_metadata$NT_test1_take<-as.numeric(cognition_metadata$NT_test1_take)
cognition_metadata$meanAL_day2to7_score<-as.numeric(cognition_metadata$meanAL_day2to7_score)
cognition_metadata$meanRL_day2to7_score<-as.numeric(cognition_metadata$meanRL_day2to7_score)
cognition_metadata$AL_slope<-as.numeric(cognition_metadata$AL_slope)
cognition_metadata$RL_slope<-as.numeric(cognition_metadata$RL_slope)

#finding the quantiles of the different groups 
#PROBLEEM SOLVING GROUPS
# Create a basic histogram to see the distribution of PS_Test1_solve
ggplot(cognition_metadata, aes(x = PS_Test1_solve)) +
  geom_histogram(binwidth = 200, fill = "steelblue", color = "black", alpha = 0.7) + # Customize bin width, color, and transparency
  labs(title = "Distribution of PS Test Solve", x = "PS Test Solve", y = "Frequency") + # Axis labels and title
  theme_minimal() # Use a minimal theme for cleaner visualization

# Calculate quantiles
PSquantiles <- quantile(cognition_metadata$PS_Test1_solve, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
PSquantiles

#PSquantiles
#    33%     66% 
# 184.95 1092.50 

# Specify breaks and labels
breaks <- c(-Inf, 185, 1093, 6000)
labels <- c("Good", "Mediocre", "Poor")

# Create 'PS_category' based on 'PS_Test1_solve'
cognition_metadata$PS_category <- cut(cognition_metadata$PS_Test1_solve, breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$PS_category<-cognition_metadata$PS_category

#see total PS_category groups for each species groups
sum_by_category <- cognition_metadata %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(PS_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)



#NOVELTY TEST GROUPS
# Create a basic histogram to see the distribution of NT_test1_take
ggplot(cognition_metadata, aes(x = NT_test1_take)) +
  geom_histogram(binwidth = 200, fill = "steelblue", color = "black", alpha = 0.7) + # Customize bin width, color, and transparency
  labs(title = "Distribution of NT Test 1 Take", x = "NT Test 1 Take", y = "Frequency") + # Axis labels and title
  theme_minimal() # Use a minimal theme for cleaner visualization

# Calculate quantiles
NTquantiles <- quantile(cognition_metadata$NT_test1_take, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
NTquantiles

#NTquantiles
#   33%    66% 
# 13.37 414.54 


# Specify breaks and labels
breaks <- c(-Inf, 14, 415, 6000)
labels <- c("Good", "Mediocre", "Poor")

# Create 'NT_category' based on 'NT_test1_take'
cognition_metadata$NT_category <- cut(cognition_metadata$NT_test1_take, breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$NT_category<-cognition_metadata$NT_category

#see total NT_category groups for each species groups
sum_by_category <- cognition_metadata %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(NT_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)


#AL TEST GROUPS
# Create a basic histogram to see the distribution of meanAL_day2to7_score
ggplot(cognition_metadata, aes(x = meanAL_day2to7_score)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "black", alpha = 0.7) + # Customize bin width, color, and transparency
  labs(title = "Distribution of mean AL", x = "AL mean", y = "Frequency") + # Axis labels and title
  theme_minimal() # Use a minimal theme for cleaner visualization

# Calculate quantiles
ALquantiles <- quantile(cognition_metadata$meanAL_day2to7_score, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
ALquantiles

#ALquantiles
#     33%      66% 
# 3.833333 6.833333 

# Specify breaks and labels
breaks <- c(-Inf, 4, 7, 15)
labels <- c("Good", "Mediocre", "Poor")

# Create 'AL_category' based on 'meanAL_day2to7_score'
cognition_metadata$AL_category <- cut(cognition_metadata$meanAL_day2to7_score, breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$AL_category<-cognition_metadata$AL_category

#see total AL_category groups for each species groups
sum_by_category <- cognition_metadata %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(AL_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)



#RL TEST GROUPS
# Create a basic histogram to see the distribution of meanRL_day2to7_score
ggplot(cognition_metadata, aes(x = meanRL_day2to7_score)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "black", alpha = 0.7) + # Customize bin width, color, and transparency
  labs(title = "Distribution of mean RL", x = "RL mean", y = "Frequency") + # Axis labels and title
  theme_minimal() # Use a minimal theme for cleaner visualization

# Calculate quantiles
RLquantiles <- quantile(cognition_metadata$meanRL_day2to7_score, probs = c(0.33, 0.66), na.rm = TRUE) 

# Print quantiles
RLquantiles

#RLquantiles
#     33%      66% 
# 4.413000 7.043333 

# Specify breaks and labels
breaks <- c(-Inf, 4, 7, 15)
labels <- c("Good", "Mediocre", "Poor")

# Create 'RL_category' based on 'meanRL_day2to7_score'
cognition_metadata$RL_category <- cut(as.numeric(cognition_metadata$meanRL_day2to7_score), breaks = breaks, labels = labels, include.lowest = TRUE)
ps_cog@sam_data$RL_category<-cognition_metadata$RL_category

#see total RL_category groups for each species groups
sum_by_category <- cognition_metadata %>%
  filter(wgs.class %in% c("CACH", "ADV.CA", "BCCH", "HYCH")) %>%
  group_by(RL_category, wgs.class) %>%
  summarise(total_individuals = n())

# Print the result
print(sum_by_category)


```





PCoA beta Diversity graphs 
```{r}

#choose which distance matrix to use
nmds_cog_jaccard <- ordinate(ps_cog, method = "NMDS", distance = jaccard)
pcoa_cog_jaccard <- ordinate(ps_cog, method = "PCoA", distance = jaccard)

nmds_cog_bray <- ordinate(ps_cog, method = "NMDS", distance = bray)
pcoa_cog_bray <- ordinate(ps_cog, method = "PCoA", distance = bray)

nmds_cog_uniunweight <- ordinate(ps_cog, method = "NMDS", distance = unifrac_unweighted)
pcoa_cog_uniunweight <- ordinate(ps_cog, method = "PCoA", distance = unifrac_unweighted)


# Extract sample data and NMDS scores
nmds_scores <- as.data.frame(nmds_cog_uniunweight$points)
pcoa_scores <- as.data.frame(pcoa_cog_uniunweight$vectors)

sample_data<-ps_cog@sam_data
# Combine sample data and NMDS scores
nmds_data <- cbind(sample_data, nmds_scores)
nmds_data <- subset(nmds_data, 
                    !is.na(wgs.class) & 
                    !is.na(nmds_data$PS_category) & 
                    !is.na(nmds_data$NT_category) & 
                    !is.na(nmds_data$AL_category) & 
                    !is.na(nmds_data$RL_category))


pcoa_data <- cbind(sample_data, pcoa_scores)
pcoa_data <- subset(pcoa_data, 
                    !is.na(wgs.class) & 
                    !is.na(pcoa_data$PS_category) & 
                    !is.na(pcoa_data$NT_category) & 
                    !is.na(pcoa_data$AL_category) & 
                    !is.na(pcoa_data$RL_category))


# Plot NMDS
nmds_plot <- ggplot(nmds_data, aes(x = MDS1, y = MDS2, color = RL_category)) +
  geom_point(size = 3) +
  labs(title = "NMDS of Beta Diversity (UniFrac Unweighted)") +
  theme_minimal() +
  stat_ellipse()


pdf("/Users/russ/Desktop/nmds_plot.pdf", width = 10, height = 10)
print(nmds_plot)
dev.off()

print(nmds_plot)


# PCoA plot
PCoA_plot<-ggplot(pcoa_data, aes(x = Axis.1, y = Axis.2, color = wgs.class, shape = wgs.class)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_minimal() +
  labs(
    x = paste0("PCoA Axis 1 (", round(pcoa_data$values$Relative_eig[1] * 100, 1), "% variance)"),
    y = paste0("PCoA Axis 2 (", round(pcoa_data$values$Relative_eig[2] * 100, 1), "% variance)"),
    color = "wgs.class",
    shape = "wgs.class"
  ) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold")
  )+
  stat_ellipse()

pdf("/Users/russ/Desktop/PCoA_plot.pdf", width = 10, height = 10)
print(PCoA_plot)
dev.off()

print(PCoA_plot)


```


DESeq work for bacteria comparisons across cog groups 
```{r}

#Deseq of cog categories (subset the *_category with the test you want to look at)

library(phyloseq)
library(DESeq2)


table(sample_data(ps_cog)$NT_category, useNA = "always")

#make an NA clean phyloseq
ps_cog_clean <- subset_samples(ps_cog, !is.na(NT_category))

#make sure the levels are correct
sample_data(ps_cog_clean)$NT_category <- factor(sample_data(ps_cog_clean)$NT_category, levels = c("Poor", "Mediocre", "Good"))

# make small numbers and get rid of low abundance bc DESeq2 doesn't handle it well
ps_cog_clean <- filter_taxa(ps_cog_clean, function(x) sum(x > 0) > (0.1 * length(x)), TRUE)

# Create DESeq2 object with RL_category as the design variable
dds <- phyloseq_to_deseq2(ps_cog_clean, ~ NT_category)

# Estimate size factors and dispersions. the "poscounts" This uses positive counts only for estimating size factors, which works much better for sparse datasets like ASV tables. this code skips zeros when estimating size factors, avoiding undefined log operations.
dds <- DESeq(dds, sfType = "poscounts")

summary(dds)

# Good vs Poor
res_good_poor <- results(dds, contrast = c("NT_category", "Good", "Poor"))

# Mediocre vs Poor
res_med_poor <- results(dds, contrast = c("NT_category", "Mediocre", "Poor"))

# Good vs Mediocre
res_good_med <- results(dds, contrast = c("NT_category", "Good", "Mediocre"))

#creates FDR threshold
res_good_poor_sig <- res_good_poor[which(res_good_poor$padj < 0.05), ]


library(ggplot2)

# Extract normalized counts
norm_counts <- counts(dds, normalized = TRUE)
top_taxa <- rownames(res_good_poor_sig)[1:min(10, nrow(res_good_poor_sig))]

# Melt and plot
library(reshape2)
# Step 1: Get normalized counts of top significant ASVs
norm_counts <- counts(dds, normalized = TRUE)
df_DESeq <- as.data.frame(t(norm_counts[top_taxa, ]))

# Step 2: Add SampleID as a column
df_DESeq$SampleID <- rownames(df_DESeq)

# Step 3: Add only the RL_category metadata
meta_df <- data.frame(sample_data(ps_cog_clean))
meta_df$SampleID <- rownames(meta_df)
meta_df <- meta_df[, c("SampleID", "NT_category")]

# Step 4: Merge
df_DESeq <- merge(df_DESeq, meta_df, by = "SampleID")
df_DESeq_melt <- melt(df_DESeq, id.vars = c("SampleID", "NT_category"))

DESeq_NT_good_poor<-ggplot(df_DESeq_melt, aes(x = NT_category, y = value, fill = NT_category)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw() +
  ylab("Normalized Abundance") +
  xlab("NT_category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Save the plot to a PDF
pdf("/Users/russ/Desktop/DESeq_NT_good_poor.pdf", width = 10, height = 10)
print(DESeq_NT_good_poor)
dev.off()


#log transforming for visualization 
log_DESeq_NT<-ggplot(df_DESeq_melt, aes(x = NT_category, y = log10(value + 1), fill = NT_category)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw() +
  ylab("Log10 Normalized Abundance") +
  xlab("NT_Category")

# Save the plot to a PDF
pdf("/Users/russ/Desktop/log_DESeq_NT.pdf", width = 10, height = 10)
print(log_DESeq_NT)
dev.off()



#adding the genus name instead of ASV#
# Pull taxonomy table
tax_DESeq <- as.data.frame(tax_table(ps_cog_clean))

# Add a column of ASV IDs to match
tax_DESeq$ASV <- rownames(tax_DESeq)

# For simplicity, keep only ASV and Genus
tax_short <- tax_DESeq[, c("ASV", "Genus")]

# Optional: Fill in missing genus names with ASV IDs
tax_short$Genus[is.na(tax_short$Genus) | tax_short$Genus == ""] <- tax_short$ASV[is.na(tax_short$Genus) | tax_short$Genus == ""]

# Create a lookup vector
asv_to_genus <- setNames(tax_short$Genus, tax_short$ASV)

df_DESeq_melt$Genus <- asv_to_genus[as.character(df_DESeq_melt$variable)]


genus_log_NT<-ggplot(df_DESeq_melt, aes(x = NT_category, y = log10(value + 1), fill = NT_category)) +
  geom_boxplot() +
  facet_wrap(~ Genus, scales = "free_y") +
  theme_bw() +
  ylab("Log10 Normalized Abundance") +
  xlab("NT_category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Save the plot to a PDF
pdf("/Users/russ/Desktop/genus_log_NT.pdf", width = 10, height = 10)
print(genus_log_NT)
dev.off()




#getting the p values 
# Convert results to dataframe if needed
res_df <- as.data.frame(res_good_poor)

# Add ASV IDs
res_df$ASV <- rownames(res_df)

# Filter for top_taxa
top_results <- res_df[rownames(res_df) %in% top_taxa, ]

# Optional: Add Genus
top_results$Genus <- asv_to_genus[top_results$ASV]

# Select key columns
top_results_clean <- top_results[, c("ASV", "Genus", "log2FoldChange", "pvalue", "padj", "baseMean")]

# Sort by adjusted p-value
top_results_clean <- top_results_clean[order(top_results_clean$padj), ]

print(top_results_clean)

write.csv(top_results_clean, "~/Desktop/top_deseq2_taxa_RL.csv", row.names = FALSE)

```

same as above but separating for the big sex differences present across the novelty test 
```{r}

# Load libraries
library(phyloseq)
library(DESeq2)
library(ggplot2)
library(reshape2)

# Extract normalized counts
norm_counts <- counts(dds, normalized = TRUE)
top_taxa <- rownames(res_good_poor_sig)[1:min(10, nrow(res_good_poor_sig))]
df_DESeq <- as.data.frame(t(norm_counts[top_taxa, ]))

# Add SampleID as a column
df_DESeq$SampleID <- rownames(df_DESeq)

# Add metadata including NT_category and Sex, remove NAs in Sex
meta_df <- data.frame(sample_data(ps_cog_clean))
meta_df$SampleID <- rownames(meta_df)
meta_df <- meta_df[, c("SampleID", "NT_category", "Sex")]
meta_df <- meta_df[!is.na(meta_df$Sex), ]

# Merge counts with metadata (automatically filters df_DESeq too)
df_DESeq <- merge(df_DESeq, meta_df, by = "SampleID")

# Melt dataframe
df_DESeq_melt <- melt(df_DESeq, id.vars = c("SampleID", "NT_category", "Sex"))

# Add Genus names
tax_DESeq <- as.data.frame(tax_table(ps_cog_clean))
tax_DESeq$ASV <- rownames(tax_DESeq)
tax_short <- tax_DESeq[, c("ASV", "Genus")]
tax_short$Genus[is.na(tax_short$Genus) | tax_short$Genus == ""] <- tax_short$ASV[is.na(tax_short$Genus) | tax_short$Genus == ""]
asv_to_genus <- setNames(tax_short$Genus, tax_short$ASV)
df_DESeq_melt$Genus <- asv_to_genus[as.character(df_DESeq_melt$variable)]

# Create plot faceted by Genus and Sex
genus_log_NT_sex <- ggplot(df_DESeq_melt, aes(x = NT_category, y = log10(value + 1), fill = NT_category)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ Genus + Sex, scales = "free_y", ncol = 4) +
  theme_bw() +
  ylab("Log10 Normalized Abundance") +
  xlab("NT_category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 8))

# Save plot
pdf("/Users/russ/Desktop/genus_log_NT_by_sex.pdf", width = 12, height = 12)
print(genus_log_NT_sex)
dev.off()


```



subsetting the alpha diversity measures into it's own dataset 
```{r}
library(vegan)
library(dplyr)
library(ggplot2)
library(phyloseq)
library(tidyverse)

# Bray-Curtis dissimilarity
bray <- phyloseq::distance(ps_cog, method = "bray")

# Jaccard dissimilarity
jaccard <- phyloseq::distance(ps_cog, method = "jaccard")

# Unweighted UniFrac
unifrac_unweighted <- phyloseq::distance(ps_cog, method = "unifrac")

# Weighted UniFrac
unifrac_weighted <- phyloseq::distance(ps_cog, method = "wunifrac")


#adding faith's phylo diversity 
# Load required library
library(picante)
library(ape)

# Root the tree
phy_tree_cog <- phy_tree(ps_cog)
rooted_tree <- root(phy_tree_cog, outgroup = phy_tree_cog$tip.label[1], resolve.root = TRUE)

# Replace the tree in the phyloseq object (optional, if you want to work with the rooted tree in ps_cog)
phy_tree(ps_cog) <- rooted_tree


# Extract tree and OTU table from ps_cog
phy_tree_cog <- phy_tree(ps_cog)
otu_cog <- as.matrix(otu_table(ps_cog))

# Calculate Faith's PD
faith_pd <- pd(otu_cog, phy_tree_cog)$PD

# Create the diversity dataframe
diversity_df <- data.frame(
  SampleID = sample_names(ps_cog),
  Shannon = estimate_richness(ps_cog, measures = "Shannon")[, "Shannon"],
  Chao = estimate_richness(ps_cog, measures = "Chao1")[, "Chao1"],
  InvSimpson = estimate_richness(ps_cog, measures = "InvSimpson")[, "InvSimpson"],
  Richness = estimate_richness(ps_cog, measures = "Observed")[, "Observed"],
  OTUs = rowSums(otu_table(ps_cog)),
  FaithPD = faith_pd
)


#making the metadata object and confirming dataframes 
meta_data<-sample_data(ps_cog)
meta_data <- data.frame(as(meta_data, "matrix"))
diversity_df<-as.data.frame(diversity_df)

# Rename seq_file to SampleID in meta_data
colnames(meta_data)[colnames(meta_data) == "seq_file"] <- "SampleID"

# Combine with metadata
diversity_df <- merge(diversity_df, meta_data, by = "SampleID")


#SEX DIFFERENCES filtering 
# Filter for all males
diversity_df_male <- diversity_df %>%
  filter(Sex == "M-DNA")
# Filter for all females
diversity_df_female <- diversity_df %>%
  filter(Sex == "F-DNA")

#sex subsetting phyloseq 
ps_cog_male <- subset_samples(ps_cog, Sex == "M-DNA")
ps_cog_female <- subset_samples(ps_cog, Sex == "F-DNA")

# Bray-Curtis dissimilarity
bray_female <- phyloseq::distance(ps_cog_female, method = "bray")
bray_male <- phyloseq::distance(ps_cog_male, method = "bray")
# Jaccard dissimilarity
jaccard_female <- phyloseq::distance(ps_cog_female, method = "jaccard")
jaccard_male <- phyloseq::distance(ps_cog_male, method = "jaccard")
# Unweighted UniFrac
unifrac_unweighted_male <- phyloseq::distance(ps_cog_male, method = "unifrac")
unifrac_unweighted_female <- phyloseq::distance(ps_cog_female, method = "unifrac")
# Weighted UniFrac
unifrac_weighted_male <- phyloseq::distance(ps_cog_male, method = "wunifrac")
unifrac_weighted_female <- phyloseq::distance(ps_cog_female, method = "wunifrac")

```


stats
```{r}
#checking for normal distribution
hist(diversity_df$Shannon, main="Histogram of Shannon Diversity", xlab="Shannon Diversity", col="lightblue")
shapiro.test(diversity_df$Shannon)

#making a no NA dataset 
df_clean <- diversity_df[!is.na(diversity_df$PS_Test1_solve), ]

library(MASS)
#checking the best fit for the family funciton of the GLMM 
fit1<-fitdistr(df_clean$PS_Test1_solve[-170], "normal")
fit2<-fitdistr(df_clean$PS_Test1_solve[-170], "lognormal")
fit3<-fitdistr(df_clean$PS_Test1_solve[-170], "exponential")
fit4<-fitdistr(df_clean$PS_Test1_solve[-170], "poisson")


AIC(fit1,fit2,fit3,fit4)


#for PS_Test1_solve the best fit was lognormal
#fit1	2	1217.935		
#fit2	2	1088.745		
#fit3	1	1119.319		
#fit4	1	185343.953	

#for NT the best fit was also lognormal 
#fit1	2	945.9896		
#fit2	2	922.0772		
#fit3	1	936.5560		
#fit4	1	31014.7289	

#for AL it was lognormal
#fit1	2	337.8788		
#fit2	2	335.7160		
#fit3	1	361.7500		
#fit4	1	Inf	

# for RL it was barely lognormal 
#fit1	2	285.8065		
#fit2	2	284.1997		
#fit3	1	343.8364		
#fit4	1	Inf	

```


glmer for ALPHA testing
```{r}
library(lme4)
library(car)

diversity_df$PS_posttrial_take<-as.numeric(diversity_df$PS_posttrial_take)
diversity_df$PS_Test1_solve<-as.numeric(diversity_df$PS_Test1_solve)
diversity_df$PS_pretrial<-as.numeric(diversity_df$PS_pretrial)
diversity_df$NT_test1_take<-as.numeric(diversity_df$NT_test1_take)
diversity_df$meanAL_day2to7_score<-as.numeric(diversity_df$meanAL_day2to7_score)
diversity_df$meanRL_day2to7_score<-as.numeric(diversity_df$meanRL_day2to7_score)

log_diversity_df<-df_clean

#log transforming all based on the best fit model 
log_diversity_df$log_PS_Test1_solve <- log(df_clean$PS_Test1_solve)
log_diversity_df$log_NT_test1_take <- log(df_clean$NT_test1_take)
log_diversity_df$log_meanAL_day2to7_score <- log(df_clean$meanAL_day2to7_score)
log_diversity_df$log_meanRL_day2to7_score <- log(df_clean$meanRL_day2to7_score)


#SHANNON DIVERSITY
#simplified version
shan.cog <- lmer(log_meanRL_day2to7_score ~ Shannon*wgs.class*Sex  + (1|Nest/Collection_year) + (1|Nest_Site), data = log_diversity_df)
summary(shan.cog)
#check the Anova 
Anova(shan.cog)

# First round of model simplification
shan.cog.new1 <- update(shan.cog, ~ . 
  - wgs.class:Sex
  - Shannon:Sex
  - Shannon:wgs.class
  - Shannon:wgs.class:Sex)

summary(shan.cog.new1)
anova(shan.cog,shan.cog.new1)
Anova(shan.cog.new1)

#CHAO DIVERSITY
#simplified version
chao.cog <- lmer(log_meanRL_day2to7_score ~ Chao*wgs.class*Sex + (1|Nest/Collection_year) + (1|Nest_Site), data = log_diversity_df)
summary(chao.cog)
#check the Anova 
Anova(chao.cog)

# First round of model simplification
chao.cog.new1 <- update(chao.cog, ~ . 
  - wgs.class:Sex
  - wgs.class:Chao
  - Chao:Sex
  - Chao:Sex:wgs.class)

summary(chao.cog.new1)
anova(chao.cog,chao.cog.new1)
Anova(chao.cog.new1)



#NUMBER OF ASVs
#simplified version
asvnum.cog <- lmer(log_meanRL_day2to7_score ~ OTUs*wgs.class*Sex  + (1|Nest/Collection_year) + (1|Nest_Site), data = log_diversity_df)
summary(asvnum.cog)
#check the Anova 
Anova(asvnum.cog)

# First round of model simplification
asvnum.cog.new1 <- update(asvnum.cog, ~ . 
  - OTUs:Sex
  - wgs.class:Sex
  - OTUs:wgs.class:Sex
  - OTUs:wgs.class)

summary(asvnum.cog.new1)
anova(asvnum.cog,asvnum.cog.new1)
Anova(asvnum.cog.new1)



#Faith's phylo diversity
#simplified version
faith.cog <- lmer(log_meanRL_day2to7_score ~ FaithPD*wgs.class*Sex  + (1|Nest/Collection_year) + (1|Nest_Site), data = log_diversity_df)
summary(faith.cog)
#check the Anova 
Anova(faith.cog)

# First round of model simplification
faith.cog.new1 <- update(faith.cog, ~ . 
  -FaithPD:Sex
  -FaithPD:wgs.class:Sex
  -wgs.class:Sex
  -FaithPD:wgs.class)

summary(faith.cog.new1)
anova(faith.cog,faith.cog.new1)
Anova(faith.cog.new1)


#####



```


PERMANOVA for BETA diversity testing 
```{r}
#perform PERMANOVA
#testing bray jaccard unifrac_weighted unifrac_unweighted
perm_cog_df <- data.frame(sample_data(ps_cog))

permanova_bray <- adonis2(bray ~ PS_category+Sex+wgs.class , 
                            data = perm_cog_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_bray)



# Impute missing values in PS_Test1_solve with the mean
diversity_clean<-diversity_df

diversity_clean$PS_Test1_solve[is.na(diversity_clean$PS_Test1_solve)] <- 
  mean(diversity_clean$PS_Test1_solve, na.rm = TRUE)
# Impute missing values in Sex with the most frequent category
diversity_clean$Sex[is.na(diversity_clean$Sex)] <- 
  names(sort(table(diversity_clean$Sex), decreasing = TRUE))[1]
# Impute missing values in wgs.class with the most frequent category
diversity_clean$wgs.class[is.na(diversity_clean$wgs.class)] <- 
  names(sort(table(diversity_clean$wgs.class), decreasing = TRUE))[1]

# Perform db-RDA
db_rda <- vegan::capscale(unifrac_weighted ~ PS_Test1_solve + Sex + wgs.class, 
                           data = diversity_clean, 
                           distance = "bray")
# Summarize results
summary(db_rda)
# ANOVA-like test to assess significance
anova(db_rda, permutations = 10000)





permanova_jaccard <- adonis2(jaccard ~ meanRL_day2to7_score + Sex + wgs.class, 
                            data = diversity_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_jaccard)


permanova_unifrac_unweighted <- adonis2(unifrac_unweighted ~ meanRL_day2to7_score + Sex + wgs.class, 
                            data = diversity_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_unifrac_unweighted)

permanova_unifrac_weighted <- adonis2(unifrac_weighted ~ meanRL_day2to7_score + Sex + wgs.class, 
                            data = diversity_df, 
                            permutations = 10000, 
                            na.action = "na.exclude", 
                            by = "margin")
print(permanova_unifrac_weighted)
```



graphs of significant interactions and aspects of alpha diversity
```{r}

otu_wgs_sex<-ggplot(diversity_df, aes(x = meanRL_day2to7_score, y = OTUs, color = wgs.class, shape = Sex)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, aes(group = wgs.class, color = wgs.class)) + 
  labs(
    x = "Reversal Learning Score", 
    y = "Number of ASVs", 
    color = "Ancestry Group", 
    shape = "Sex"
  ) +
  theme_minimal() +
  theme_cowplot(12)

pdf("/Users/russ/Desktop/otu_wgs_sex.pdf", width = 10, height = 10)
print(otu_wgs_sex)
dev.off()

```





